Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-28 Manuscript/manuscript.Rmd
edits

where $C_{min}$ denotes the set of TAD boundary regions and $C_{maj}$ denotes the set non-TAD boundary regions. CI of heavily imbalanced classes approaches 0, while CI for balanced classes approaches 0.50. To assess the effect of CI on predictive performance, we first used the data with **o Sampling**, that is, the majority and minority classes were used as is. We evaluated three re-sampling techniques: **Random Under-Sampling (RUS)** - ll of the data points from the minority class of the training set were used. Sampling without replacement was used to obtain the same number of data points from the majority class.

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. This can contribute to high collinearity which can result in overfitting a predictive model. To measure the pairwise correlation of each cell line specific annotation, we used the Jaccard Index. The Jaccard Index measures the ratio of the number of intersecting base pairs between two sets to the number of base pairs in the union, subtracted by the intersection [@favorov2012exploring]. The quantity ranges from 0.0 to 1.0, where 0.0 represents no overlap (no correlation) and 1.0 represent complete overlap (complete correlation).

In order to suitably reduce the feature space and prevent overfitting, we implemented recursive feature elimination to estimate the optimal number of necessary features to include in the model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. Given that our approach is chromosome specific, we aggregated the predictive importance of the union of the optimal set of features across chromosome using the mean decrease in node impurity. We then performed hierarchical clustering to obtained the most common and top ranked annotations responsible for TAD boundary prediction.

@@

Once we established our optimally built and feature reduced random forest classifier, we aimed to leverage the enrichment of known annotations to predict TAD boundaries at base pair resolution. We established a *predicted TAD boundary region* as a collection of adjacent base pairs that were given a probability of 1 by our model in being a predicted TAD boundary. To guard against inflating the number of predicted TAD boundaries, we clustered the distances between the midpoints of the *predicted TAD boundary region*. We then performed partitioning around mediods (PAM) to determine the *predicted TAD boundary region* that was most representative of the cluster. Lastly, the midpoint, in this case a particular basepair, of the *predicted TAD boundary region* identified as the medoid was defined as a *predicted TAD boundary*.

>>>>

where $C_{min}$ denotes the set of TAD boundary regions and $C_{maj}$ denotes the set non-TAD boundary regions. CI of heavily imbalanced classes approaches 0, while CI for balanced classes approaches 0.50. To assess the effect of CI on predictive performance, we first used the data with **No Sampling**, that is, the majority and minority classes were used as is. We evaluated three re-sampling techniques: **Random Under-Sampling (RUS)** - all of the data points from the minority class of the training set were used. Sampling without replacement was used to obtain the same number of data points from the majority class.

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. <!--This can contribute to high collinearity which can result in overfitting a predictive model.--> To measure the pairwise correlation of each cell line specific annotation, we used the Jaccard Index. The Jaccard Index measures the ratio of the number of intersecting base pairs between two sets to the number of base pairs in the union, subtracted by the intersection [@favorov2012exploring]. The quantity ranges from 0 to 1, where 0 represents no overlap (no correlation) and 1 represent complete overlap (complete correlation).

In order to suitably reduce the feature space and prevent overfitting, we implemented recursive feature elimination to estimate the optimal number of necessary features to include in the model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosome using the mean decrease in node impurity. Hierarchical clustering was performed to obtained the most common and top ranked annotations responsible for TAD boundary prediction.

@@

By leveraging the enrichment of known annotations in our model, we established a *predicted TAD boundary region* as a collection of adjacent base pairs that were given a probability of 1 for being a potentially predicted TAD boundary. To guard against inflating the number of predicted TAD boundaries, we clustered the distances between the midpoints of the *predicted TAD boundary regions*. We then performed partitioning around mediods (PAM) to determine the *predicted TAD boundary region* that was most representative of the cluster. The midpoint, in this case a particular basepair, of the *predicted TAD boundary region* identified as the medoid was defined as a final *predicted TAD boundary*. This process was repeated throughout the chromosome-specific sections of the linear genome.
